Meta-analytic approaches and effect sizes to account for 'nuisance heterogenity' in comparative physiology

### 2021-07-21 MS/ms.Rmd
Patrices comments for MS included

```{r, figsurv, fig.cap="**Summary of meta-analytic practices in comparative physiology.** (A) Different types of effect measures commonly used in meta-analyses and their relative frequency of use. (B) How many meta-analyses where ‘nuisance heterogenity’ have the potential to impact effect sizes, and the proportion of studies explicitly dealing with nuisance heterogenity. (C) Frequency of analytical approaches used to analyse effect measures. Note that only ‘Weighted Regression’, ‘Random Effects Models’, ‘Multilevel Metaregression Models’ and ‘Multilevel Meta-analysis Models’ are weighted models. The results of this survey were drawn from 62 studies in comparative physiology."}
p1 = dat %>% dplyr::group_by(effect_size_type) %>% janitor::tabyl(effect_size_type)  %>% 
  dplyr::summarise(SMD = dplyr::if_else(grepl("SMD", effect_size_type), n, 0),
                  lnRR = dplyr::if_else(grepl("Response Ratio", effect_size_type), n, 0),
                    Zr = dplyr::if_else(grepl("correlation", effect_size_type), n, 0),
                 Other = dplyr::if_else(grepl("other|Q10", effect_size_type), n, 0),
                 Means = dplyr::if_else(grepl("mean/proportion/raw", effect_size_type), n, 0),
                 lnCVR = dplyr::if_else(grepl("lnCVR", effect_size_type), n, 0),
                 Slope = dplyr::if_else(grepl("slope", effect_size_type), n, 0)) %>% 
  colSums(.) %>% data.frame(effect = names(.), n = .) %>% mutate(prop = n / sum(n)) %>% arrange(effect) %>% 
  mutate(effect = fct_relevel(effect, c("Other", "Zr", "lnRR", "SMD", "lnCVR", "Slope", "Means"))) %>% 
  ggplot() + tm + 
  geom_col(aes(x = effect, y = prop), width = 0.5, fill = "black", colour = text.colour, size = 0.5) +
  scale_y_continuous(expand=c(0,0), limits = c(0,0.35)) +
  labs(x = "Effect Size",
       y = "Frequency") +
  coord_flip()

p2 = dat %>% mutate(cont_moderator_variable_modelled = if_else(cont_moderator_variable_modelled == "", "N/A", cont_moderator_variable_modelled)) %>% 
  ggplot() + tm +
  geom_bar(aes(x = cont_moderator_variable,
               fill = cont_moderator_variable_modelled,
               y = ..count../sum(..count..)), width = 0.5,
           colour = text.colour, size = 0.5) +
  scale_y_continuous(expand=c(0,0), limits = c(0,1)) +
  scale_fill_viridis_d() +
  theme(legend.position = c(0.50, 0.8)) + 
  labs(x = "Nuisance Heterogeneity?",
       y = "Frequency",
       fill = "Modeled Nuisance\nHeterogeneity?")

@@

A common goal of meta-analysis is to obtain an overall meta-analytic mean effect size estimate and some form of uncertainty around that mean, like a 95% confidence interval (CI). Including a prediction interval (PI) can also be quite useful (see below for why) [@Nakagawa2021c]. Estimating an overall mean and 95% CI is generally done with a simple random effects or multilevel meta-analytic model that accounts for effect size sampling error [@Nakagawa2017; @NakagawaSantos2012]. More often than not, meta-analyses in comparative physiology will have complex hierarchical structure. For example, multiple effect sizes may be taken from a single study on different traits measured on the same individuals, or come from many different species that share an evolutionary history [@Cinar2021; @Noble2017]. As such, it is far more likely that comparative physiologists will need to apply a multi-level meta-analytic model to control for the various sources of non-independence, and whenever possible they should [@Cinar2021; @NakagawaSantos2012; @Song2021; @Nakagawa2021; @Noble2017]. Multilevel models are also extremely useful in describing the various sources contributing to effect size variance (a feature we describe below). For the purpose of our discussion, we will assume the meta-analyst needs to use more advanced multilevel models. Having said that, it may be sufficient for a meta-analysis to make use of traditional random effects models. We refer readers to excellent reviews and books that overview the distinctions between common (fixed) effect, random effect and multilevel models and when one might want (or not want) to apply these models [@Nakagawa2017; @NakagawaSantos2012; @Koricheva2013; @Borenstein2009; @Schmid2021].

Assume that we are interested in understanding the overall effect of, say, stress (i.e., corticosterone - CORT) on reproduction [e.g., see @Podmola2018]. We might extract data from studies manipulating corticosterone in bird eggs. Of course, these studies are diverse, apply different dosages of corticosterone and are working across many species. If the studies are experimental in nature then a standardised effect size, such as $ln_{RR}$ or Hedges' *d*, might be applicable. We then might fit an overall multilevel meta-analytic model as follows: 
$$
\begin{split}
Y_{i} = \beta_{0} + a_{k[i]} + sp_{k[i]} + s_{j[i]} + e_{i} + m_{i} \\
a_{k[i]}\sim N(0, \sigma^{2}_{phylogeny}\textbf{A}), \\
sp_{k[i]}\sim N(0, \sigma^{2}_{species}\textbf{I}), \\
s_{j[i]}\sim N(0, \sigma^{2}_{study}\textbf{I}), \\
e_{i}\sim N(0, \sigma^{2}_{residual}\textbf{I}), \\
m_{i} \sim N(0, v_{i}\textbf{I}) \\
\end{split}
$$

where $Y_{i}$ is the *i*th effect size (*i* = 1 ... $N_{ES}$, the number of effect sizes), $\beta_{0}$ is the overall meta-analytic mean, $a_{k[i]}$ is the phylogenetic effect (a random effect) for species *k* applied to effect size *i* [i.e., species effects because of shared evolutionary history, @Cinar2021]. Phylogenetic effects are assumed to be normally distributed deviates sampled from a distribution with a mean of 0 and variance $\sigma^{2}_{phylogeny}$ [i.e., *N*$(0, \sigma^{2}_{phylogeny}\textbf{A})$], where $\textbf{A}$ is a phylogenetic correlation matrix derived from a phylogenetic tree. $sp_{k[i]}$ is an additional species-specific effect for the *k*th species applied to effect size *i* [i.e., these are species-specific effects because of shared ecology and other factors, @Cinar2021], $s_{j[i]}$ is the study-specific effect for the *j*th study applied to effect size *i*, $e_{i}$ is the effect size-specific effect (or within study effect, or residuals) for the *i*th effect size, and $m_{i}$ is the sampling variance effect around the *i*th effect as a result of variation in precision among each effect sizes, $v_{i}$, which is known as the sampling variance for effect. This model is a 'weighted' multilevel meta-analysis model because the estimation of the overall meta-analytic mean, $\beta_{0}$, is weighted by the inverse sampling variance of each effect size. Importantly, if we have effect sizes that are derived from the *same* sample of animals (or plants) (e.g., may be because many traits are measured on the same sets of animals) then we need to remove this dependency from the calculation of each $v_{i}$; we can then model the known sampling covariance between effect sizes by modifying $\textbf{I}$ to include off-diagonals that account for this covariance [See Box 1 and @Noble2017]. The fact that we can obtain an overall estimate of sampling variance (i.e., $\sigma^2_{m}$, see below for how it is defined) is very important because it provides a way for us to understand just how much variation in effects are the result of sampling differences across studies versus other, potentially important sources of variation such as differences in the methods applied, species differences or even real biological effects that impact the effect size a study is likely to observe [@Nakagawa2017; @NakagawaSantos2012]. This is formalised in what we call 'heterogeneity analysis' which we describe below. Importantly, this model provides us with the answer to our overall question: Across all the species and studies manipulating egg corticosterone, how much of an overall effect does corticosterone have on reproduction relative to control conditions?

Of course, the overall effect does not provide us a way to interpret its magnitude in the context of how much CORT has been applied in the treatments used in the sample of studies. The most we can say about it, is that the effect is small, moderate or large out of the sample of effect sizes extracted from this set of studies. In the next section, we show how to 'correct' the overall effect size so that its interpretation is more meaningful.

@@

where $\sum \beta_{j}x_{j}$ is simply the sum of all effects for all moderators (fixed effects), $x_{j}$. All other notation is the same as described above. Importantly, the variable $x_{j}$ could be any moderator collected from studies, including dosage, temperature, salinity, or pH. Some moderators will be at the level of the study (e.g., was the study experimental or observational in nature) but others could be taken at the effect size level (i.e., was the effect derived using a temperature of 23°C or 35°C). By including these moderators in a meta-regression model, the interpretation of the overall meta-analytic mean, $\beta_{0}$, changes. For example, assume we collected the change in CORT dosage relative to a control group (i.e., in uL / g) for our meta-analysis on egg CORT effects on bird reproduction. If the untransformed dosage, $x_{j}$, is included in the model then the overall meta-analytic mean estimate, $\beta_{0}$, would be the mean effect of CORT relative to a control treatment, on reproduction when the dosage difference between the treatment and control group is 0 uL /g. Of course, this interpretation does not make much sense.

We can, however, apply some simple transformations to make the overall mean estimate more intuitive to understand with the added benefit that it can help improve the interpretation of non-linear parameters estimated by the model [@Schielzeth2010; @GelmanHill2007]. One simple transformation would be to centre the variable $x_{j}$, by subtracting each value from the mean to create a new input variable (i.e., $c_{i}=x_{i}-\bar{x}$, where the subscript *i* denotes each individual effect size dosage). The new variable, $c_{j}$, is now centered on the mean and can replace $x_{j}$ in the meta-regression model. When the model is refit with $c_{j}$ the estimated $\beta_{0}$ can now be more intuitively interpreted as the overall mean effect of CORT on reproduction at an average dosage difference between treatment and control groups. Importantly, by centering, we retain the original units (uL / g) of the variable, which can be very useful to allow for a more general interpretation of the effect magnitude (e.g., when the dosage difference is 10 uL /g). Given it is very common to estimate and compare overall meta-analytic means (i.e., $B_{0}$) in meta-analyses this centering can be particularly useful in making these means more interpretable. Mean centering can even be extended to include other 'nuisance' variables that might creep into the dataset; the same centering can be done with a second variable, and the overall mean adjusted can be interpreted in the context of this new variable in addition to dosage. Of course, if one is not interested in the magnitude of the effect, we could also change our question and simply model mean reproductive output as a function of dosage (often what are referred to as 'arm-based' meta-analytic models). Here, one can model the mean for each group (i.e., control and treatment), and account for the known sampling variance associated with each group mean. Dosage, or other continuous variables like temperature, can then be modeled (accounting for sampling variance) to understand how mean reproductive output changes. It is important to recognise that all means need to be comparable and that these models are more complex to fit. While they can be equivalent to 'contrast-based' models, they require complex interactions to be estimated [@Nakagawa2015].

Centering moderator variables, like temperature and dosage, not only provides a way to make the overall mean effect more interpretable in the face of treatment heterogeneity, but it also allows more complex relationships to be fit without compromising interpretations, such as when one includes non-linear parameters (e.g., polynomials) and interactions with other variables. In other words, linear / main effects can still be interpreted normally in the presence of non-linear / interactions fit in the model [@Schielzeth2010; @GelmanHill2007]. We provide examples in the online supplementary material (with R code) to show readers how the models we describe above can be fit, and how, meta-analytic means can be adjusted for treatment heterogeneity.

@@

As we have already emphasised, quantifying effect size heterogeneity (i.e., effect size variance) provides the impetus for exploring hypotheses that are thought to explain why effects might vary within and across studies, while also providing context for interpreting the generality of effects. There are a number of measures commonly used to quantify effect size heterogeneity [@Borenstein2019], however, in ecological and evolutionary meta-analysis, heterogeneity is often calculated and reported as $I_{total}^2$ as follows (if using a multilevel meta-analytic model as we describe above):

@@

Quantifying heterogeneity using $I_{total}^2$ provides a formal way to assess the relative amount of variation that is the result of 'true' biological or methodological differences as opposed to simply chance (i.e., sampling variance) [@NakagawaSantos2012; @Higgins2002]. In other words, $I_{total}^2$ can be thought of as the percentage of variance between effect sizes after removing the effects of sampling error [@Higgins2002]. We can only calculate this metric if we have an estimate of the sampling variance for a given effect measure because only then can we quantify the proportion of variation that is the result of sampling variability. @Senior2016 have shown that total heterogeneity is often extremely high in ecological and evolutionary meta-analysis (\~91%), suggesting that effect measures vary widely across and within studies. Some of the variation is clearly the result of working with different species or strains, different traits and different outcome measures, however, some of this variation also arises because of methodological differences across studies that might be related to the dosages or temperatures used for treatments.

Formally quantifying and presenting heterogeneity estimates from meta-analytic models can provide a way to understand the major drivers of effect size variation by allowing researchers to produce various $I^2$ estimates which can be compared within and across studies. For example, we could fit the multilevel model described above to understand what proportion of variation is the result of, say, 'study' effects following @NakagawaSantos2012: $I^2_{study} = \frac{\sigma^2_{study}}{\sigma^2_{total}}$ where $\sigma^2_{study}$ is the study specific variance.

Many readers will notice the similarities between $I^2$ and $R^2$ [@Nakgawa2013; @Nakagawa2010; @Nakagawa2017]. If we want to formally assess how much effect size variation is the result of 'nuisance' variation, such as temperature and dosage differences applied across studies, then we could fit our multilevel meta-regression model, including dosage or temperature (linear, quadratic or even more complex fits), and estimate how much variation is explained by these fixed effects following @Nakgawa2013.

@@

Note that this formula does not include $\sigma^2_{m}$ as sampling error variance is assumed to be known in meta-analysis as explained above. When including continuous moderators, such as dosage, temperature, salinity, and even pH, $R^2_{marginal}$ provides a formal means to assess just how much variation in effects is the result of 'nuisance' heterogeneity at the effect size level. Other biological moderators of interest could also be included as fixed effects and $R^2_{marginal}$ calculated for each independently or all together. Importantly, moderators could be those that explain variation at the effect size level (i.e., dosage, temperature) or they could explain variation at the study or species level (e.g., reproductive mode, endothermy etc). $R^2_{marginal}$, or how we like to think of it in this context as $R^2_{'nuisance'}$, maybe a useful measure to help readers understand just how much effect variation can result from the specific treatment application. If the heterogeneity is high as a result of, say, dosages differences, then it is clear that the choice of dosage will have a critical impact on the effect of interest. Comparative physiologists interested in implementing these calculations can do so using our packages `orchaRd` [@Nakagawa2021c] or `metaAidR` (<https://github.com/daniel1noble/metaAidR>).

Heterogeneity described using $I^2$ measures can be useful for understanding the relative contributions of different factors to effect size variation, however, prediction intervals might be more appropriate in many cases [@Nakagawa2021c; @Borenstein2019]. Unlike $I^2$, prediction intervals (PI's) are probably more meaningful in the context of meta-analysis because they explicitly incorporate measures of dispersion around a mean effect size providing information about the likely effect size one can expect if we were to randomly sample a new population. For example, assume that we were interested in the overall impact of salinity stress in freshwater fish. To tackle this question, we collected experimental studies measuring fish swim performance under high salinity treatments (\~15-20ppm Na Cl) compared to freshwater (\~0ppm Na Cl) controls; using SMD to quantify the magnitude of effect salinity had on swim performance. We conduct a meta-analysis and estimate an overall SMD of 0.50 with a prediction interval of 0.05 to 0.85. The PI indicates that effects can vary widely from as low as an SMD of 0.05 to as high as an SMD 0.95, depending on the population. In addition, if we were to repeat a study we would expect a new effect to fall within the range of 0.05 and 0.85, 95% of the time [@Borenstein2019; @Nakagawa2021c]. This is a very intuitive interpretation of how variable effects are, yet PIs are often not reported in meta-analyses [@Nakagawa2021c], and we encourage more meta-analysts to report these.

@@

Comparative physiologists are interested in meta-analysing effects across a diversity of species, experimental approaches / designs and environments. Importantly, sampling variances for many 'effect measures' commonly used by comparative physiologists (e.g., CTmax, Slopes etc) can be formalised and meta-analytic models applied; even though it is still common to see studies masquerading as meta-analyses despite not fitting meta-analytic models. While an overall meta-analytic mean may be of interest, more often than not, understanding the drivers of effect size heterogeneity will be the primary interest. However, obvious drivers of effect size heterogeneity, such as temperature and dosage differences across studies, may not really be of interest. Comparative physiologists are likely to want an effect size or analytic approach that allows them to remove these 'nuisances'. Here, we provide a set of tools to deal with 'nuisance heterogeneity' by 'reinventing' standard effect sizes or using meta-regression models. Of course, 'nuisance heterogeneity' might in of itself be the main interest, in which case standard meta-regression approaches will work sufficiently. However, in many cases, comparative physiologists might simply want to focus on bigger, more interesting biological drivers of effect size direction and magnitude. Estimating complex meta-regression models with a multitude of fixed effects might result, however, in more challenging model to interpret As such, a combined approach, where some of the known heterogeneity is dealt with by deriving a variant effect size and associated sampling variance, and other biologically interesting moderators are fit in a meta-regression model, might be desired. Indeed, this may even be a necessity to simplify the model. We hope the set of tools we describe provide clarification around different sources of heterogeneity; providing new ways to help comparative physiologists communicate effect measures more richly to guide our understanding and decisions around pressing global challenges.

@@

We would like to thank the Editorial board (Craig Franklin, Michaela Handel and Charlotte Rutledge) for inviting us to contribute to this special issue along with feedback on previous drafts. We would also like to than Megan Head for comments on a previous draft. DWAN, SN, ML and REO were supported by ARC Discovery Projects (DP210101152 to DWAN and DP200100367 to SN, ML and DWAN).

>>>>

```{r, figsurv, fig.wid = 16.833333, fig.height=6.611111,fig.cap="**Summary of meta-analytic practices in comparative physiology.** (A) Different types of effect measures commonly used in meta-analyses and their relative frequency of use. (B) How many meta-analyses where ‘nuisance heterogenity’ have the potential to impact effect sizes, and the proportion of studies explicitly dealing with nuisance heterogenity. (C) Frequency of analytical approaches used to analyse effect measures. Note that only ‘Weighted Regression’, ‘Random Effects Models’, ‘Multilevel Metaregression Models’ and ‘Multilevel Meta-analysis Models’ are weighted models. The results of this survey were drawn from 62 studies in comparative physiology."}
p1 = dat %>% dplyr::group_by(effect_size_type) %>% janitor::tabyl(effect_size_type)  %>% 
  dplyr::summarise(SMD = dplyr::if_else(grepl("SMD", effect_size_type), n, 0),
                  lnRR = dplyr::if_else(grepl("Response Ratio", effect_size_type), n, 0),
                    Zr = dplyr::if_else(grepl("correlation", effect_size_type), n, 0),
                 Other = dplyr::if_else(grepl("other|Q10", effect_size_type), n, 0),
                 Means = dplyr::if_else(grepl("mean/proportion/raw", effect_size_type), n, 0),
                 lnCVR = dplyr::if_else(grepl("lnCVR", effect_size_type), n, 0),
                 Slope = dplyr::if_else(grepl("slope", effect_size_type), n, 0)) %>% 
  colSums(.) %>% data.frame(effect = names(.), n = .) %>% mutate(prop = n / sum(n)) %>% arrange(effect) %>% 
  mutate(effect = fct_relevel(effect, c("Other", "Zr", "lnRR", "SMD", "lnCVR", "Slope", "Means"))) %>% 
  ggplot() + tm + 
  geom_col(aes(x = effect, y = prop), width = 0.5, fill = "black", colour = text.colour, size = 0.5) +
  scale_y_continuous(expand=c(0,0), limits = c(0,0.35)) +
  labs(x = "Effect Size",
       y = "Frequency") +
  coord_flip()

p2 = dat %>% mutate(cont_moderator_variable_modelled = if_else(cont_moderator_variable_modelled == "", "N/A", cont_moderator_variable_modelled)) %>% 
  ggplot() + tm +
  geom_bar(aes(x = cont_moderator_variable,
               fill = cont_moderator_variable_modelled,
               y = ..count../sum(..count..)), width = 0.5,
           colour = text.colour, size = 0.5) +
  scale_y_continuous(expand=c(0,0), limits = c(0,1)) +
  scale_fill_viridis_d() +
  theme(legend.position = c(0.45, 0.8)) + 
  labs(x = "Nuisance Heterogeneity?",
       y = "Frequency",
       fill = "Modeled Nuisance\nHeterogeneity?")

@@

A common goal of meta-analysis is to obtain an overall meta-analytic mean effect size estimate and some form of uncertainty around that mean, like a 95% confidence interval (CI). Including a prediction interval (PI) can also be useful (discussed below) [@Nakagawa2021c]. Estimating an overall mean and 95% CI is generally done with a simple random effects or multilevel meta-analytic model that accounts for effect size sampling error [@Nakagawa2017; @NakagawaSantos2012]. More often than not, meta-analyses in comparative physiology will have complex hierarchical structure. For example, multiple effect sizes may be taken from a single study on traits measured on the same individuals, or come from many different species that share an evolutionary history [@Cinar2021; @Noble2017]. As such, it is far more likely that comparative physiologists will need to apply a multilevel meta-analytic model to control for the various sources of non-independence, and whenever possible, they should [@Cinar2021; @NakagawaSantos2012; @Song2021; @Nakagawa2021; @Noble2017]. Multilevel models are also extremely useful in describing the various sources contributing to effect size variance (a feature we describe below). For the purpose of our discussion, we will assume one needs to use more advanced multilevel models. Having said that, in some instances, it may be sufficient for a meta-analysis to make use of traditional random effects models. We refer readers to excellent reviews and books that overview the distinctions between common (fixed) effect, random effect and multilevel models and when one might want (or not want) to apply these models [@Nakagawa2017; @NakagawaSantos2012; @Koricheva2013; @Borenstein2009; @Schmid2021].

Assume that we are interested in understanding the overall effect of, say, stress (i.e., corticosterone - CORT) on reproduction [e.g., see @Podmola2018]. We might extract data from studies manipulating corticosterone in bird eggs. Of course, these studies are diverse, spanning many species and applying different dosages of corticosterone and are working across many species. If the studies are experimental in nature, then a standardised effect size, such as $ln_{RR}$ or SMD, might be applicable. We then might fit an overall multilevel meta-analytic model as follows: 
$$
\begin{split}
Y_{i} = \beta_{0} + a_{k[i]} + sp_{k[i]} + s_{j[i]} + e_{i} + m_{i} \\
a_{k[i]}\sim N(0, \sigma^{2}_{phylogeny}\textbf{A}), \\
sp_{k[i]}\sim N(0, \sigma^{2}_{species}\textbf{I}), \\
s_{j[i]}\sim N(0, \sigma^{2}_{study}\textbf{I}), \\
e_{i}\sim N(0, \sigma^{2}_{residual}\textbf{I}), \\
m_{i} \sim N(0, v_{i}\textbf{I}) \\
\end{split}
$$

where $Y_{i}$ is the *i*th effect size (*i* = 1 ... $N_{ES}$, the number of effect sizes), $\beta_{0}$ is the overall meta-analytic mean, $a_{k[i]}$ is the phylogenetic effect (a random effect) for species *k* applied to effect size *i* [i.e., species effects because of shared evolutionary history, @Cinar2021]. Phylogenetic effects are assumed to be normally distributed deviates sampled from a distribution with a mean of 0 and variance $\sigma^{2}_{phylogeny}$ [i.e., *N*$(0, \sigma^{2}_{phylogeny}\textbf{A})$], where $\textbf{A}$ is a phylogenetic correlation matrix derived from a phylogenetic tree. $sp_{k[i]}$ is an additional species-specific effect for the *k*th species applied to effect size *i* [i.e., these are species-specific effects because of shared ecology and other factors, @Cinar2021], $s_{j[i]}$ is the study-specific effect for the *j*th study applied to effect size *i*, $e_{i}$ is the effect size-specific effect (or within study effect, or residuals) for the *i*th effect size, and $m_{i}$ is the sampling variance effect around the *i*th effect as a result of variation in precision among each effect sizes, $v_{i}$, which is known as the sampling variance for the effect. This model is a 'weighted' multilevel meta-analysis model because the estimation of the overall meta-analytic mean, $\beta_{0}$, is partially weighted by the inverse sampling variance of each effect size. Importantly, if we have effect sizes that are derived from the *same* sample of organisms (e.g., because traits are measured on the same sets of individuals) then we need to remove this dependency from the calculation of each $v_{i}$; we can then model the known sampling covariance between effect sizes by modifying $\textbf{I}$ to include off-diagonals that account for this covariance [See Box 1 and @Noble2017]. The fact that we can obtain an overall estimate of sampling variance (i.e., $\sigma^2_{m}$, see below for how it is defined) is very important because it provides a way for us to understand just how much variation in effects are the result of sampling differences across studies versus other, potentially important sources of variation such as differences in the methods applied, species differences or even real biological effects that impact the effect size a study is likely to observe [@Nakagawa2017; @NakagawaSantos2012]. This is formalised in what we call 'heterogeneity analysis' which we describe below. Importantly, this model provides us with the answer to our overall question: Across all the species and studies manipulating egg corticosterone, how much of an overall effect does corticosterone have on reproduction relative to control conditions?

Of course, the overall effect does not provide us with a way to interpret its magnitude in the context of how much CORT has been applied in the treatments used in the sample of studies. The most we can say about it, is that the effect is small, moderate or large out of the sample of effect sizes extracted from this set of studies. In the next section, we show how to 'correct' the overall effect size so that its interpretation is more meaningful.

@@

where $\sum \beta_{j}x_{j}$ is simply the sum of all effects for all moderators (fixed effects), $x_{j}$. All other notation is the same as described above. Importantly, the variable $x_{j}$ could be any moderator collected from studies, including dosage, temperature, salinity, or pH. Some moderators will be at the level of the study (e.g., was the study experimental or observational in nature) but others could be taken at the effect size level (i.e., was the effect derived using a temperature of 23°C or 35°C). By including these moderators in a meta-regression model, the interpretation of the overall meta-analytic mean, $\beta_{0}$, changes. For example, assume we collected the change in CORT dosage relative to a control group (i.e., in µL / g) for our meta-analysis on egg CORT effects on bird reproduction. If the untransformed dosage, $x_{j}$, is included in the model then the overall meta-analytic mean estimate, $\beta_{0}$, would be the mean effect of CORT on reproduction when the dosage difference between the treatment and control group is 0 µL /g. Of course, this interpretation does not make much sense.

We can, however, apply some simple transformations to make the overall mean estimate more intuitive while also improving the interpretation of non-linear parameters and interactions estimated from the model [@Schielzeth2010; @GelmanHill2007]. One simple transformation would be to centre the variable $x_{j}$, by subtracting each value from the mean to create a new input variable (i.e., $c_{i}=x_{i}-\bar{x}$, where the subscript *i* denotes each individual effect size dosage). The new variable, $c_{j}$, is now centered on the mean and can replace $x_{j}$ in the meta-regression model. When the model is refit with $c_{j}$, the estimated $\beta_{0}$ can now be more intuitively interpreted as the overall mean effect of CORT on reproduction at an average dosage difference between treatment and control groups. Importantly, by centering, we retain the original units (µL / g) of the variable, which can be very useful to ease the interpretation of the effect magnitude (e.g., when the dosage difference is 10 µL /g). Given it is very common to estimate and compare overall meta-analytic means (i.e., $B_{0}$) in meta-analyses this centering can be particularly useful in making these means more interpretable. Mean centering can even be extended to include other 'nuisance' variables that might creep into the dataset; the same centering can be done with a second variable, and the overall mean adjusted and interpreted in the context of this new variable along with dosage. Of course, if one is not interested in the magnitude of the effect, we can simply model mean reproductive output as a function of dosage (often what are referred to as 'arm-based' meta-analytic models). Here, one can model the mean for each group (i.e., control and treatment), and account for the known sampling variance associated with each group mean. Dosage, or other continuous variables like temperature, can then be modeled (accounting for sampling variance) to understand how mean reproductive output varies. It is important to recognise that all means need to be comparable and that these models are more complex to fit. While they can be equivalent to 'contrast-based' models, they require complex interactions to be estimated [@Nakagawa2015].

Centering moderator variables, like temperature and dosage, not only provides a way to make the overall mean effect more interpretable in the face of treatment heterogeneity, but it also allows more complex relationships to be fit without compromising interpretations, such as when one includes non-linear parameters (e.g., polynomials) and interactions with other variables. In other words, linear / main effects can still be interpreted normally in the presence of non-linear / interactions fit in the model [@Schielzeth2010; @GelmanHill2007]. We provide examples in the online supplementary material (with R code) to show readers how the models we describe above can be fit, and how meta-analytic means can be adjusted for treatment heterogeneity.

@@

As we have already emphasised, quantifying effect size heterogeneity provides the impetus for exploring why effects might vary within and across studies, while also providing context for interpreting the generality of effects. There are a number of measures commonly used to quantify effect size heterogeneity [@Borenstein2019], however, in ecological and evolutionary meta-analysis, heterogeneity is often calculated and reported as $I_{total}^2$ as follows (if using a multilevel meta-analytic model as we describe above):

@@

Quantifying heterogeneity using $I_{total}^2$ provides a formal way to assess the relative amount of variation that is the result of 'true' biological or methodological differences as opposed to simply chance (i.e., sampling variance) [@NakagawaSantos2012; @Higgins2002]. In other words, $I_{total}^2$ is the percentage of variance between effect sizes after removing the effects of sampling error [@Higgins2002]. Note that we can only calculate this metric if we have an estimate of the sampling variance for a given effect measure because only then can we quantify the proportion of variation that is the result of sampling variability. @Senior2016 have shown that total heterogeneity is often extremely high in ecological and evolutionary meta-analysis (\~91%), suggesting that effect measures vary widely within and across studies. Some of the variation is clearly the result of working with different species and/or strains, or measuring different traits and outcome measures. However, some of this variation also arises because of methodological differences across studies that might be related to the dosages or temperatures used for treatments.

Formally quantifying and presenting heterogeneity estimates from meta-analytic models can provide a way to understand the major drivers of effect size variation by producing various $I^2$ estimates that can be compared within and across studies. For example, we could fit the multilevel model described above to understand what proportion of variation is the result of, say, 'study' effects following @NakagawaSantos2012: $I^2_{study} = \frac{\sigma^2_{study}}{\sigma^2_{total}}$ where $\sigma^2_{study}$ is the study specific variance.

Some readers may notice the similarities between $I^2$ and $R^2$ [@Nakgawa2013; @Nakagawa2010; @Nakagawa2017]. If we want to formally assess how much effect size variation is the result of 'nuisance heterogeneity' (e.g.,  temperature and dosage differences applied across studies) then we could fit our multilevel meta-regression model, including dosage or temperature (linear, quadratic or even more complex fits), and estimate how much variation is explained by these fixed effects following @Nakgawa2013.

@@

Note that this formula does not include $\sigma^2_{m}$ as sampling error variance is assumed to be known in meta-analysis, as explained above. When including continuous moderators, such as dosage, temperature, salinity, and even pH, $R^2_{marginal}$ provides a formal means to assess just how much variation in effects is the result of 'nuisance' heterogeneity at the effect size level. Other biological moderators of interest could also be included as fixed effects and $R^2_{marginal}$ can be calculated for each independently, or all together. Importantly, moderators could be those that explain variation at the effect size level (i.e., dosage, temperature) or variation at the study or species level (e.g., reproductive mode, endothermy etc). $R^2_{marginal}$, or how we like to think of it in this context as $R^2_{'nuisance'}$, maybe a useful measure to help readers understand just how much effect variation can result from the specific treatment application. If the heterogeneity is high as a result of, say, dosages differences, then it is clear that the choice of dosage will have a critical impact on the effect of interest. Comparative physiologists interested in implementing these calculations can do so using our packages `orchaRd` [@Nakagawa2021c] or `metaAidR` (<https://github.com/daniel1noble/metaAidR>).

Heterogeneity described using $I^2$ measures can be useful for understanding the relative contributions of different factors to effect size variation, however, prediction intervals might be more appropriate in many cases [@Nakagawa2021c; @Borenstein2019]. Unlike $I^2$, prediction intervals (PI's) are probably more meaningful in the context of meta-analysis because they explicitly incorporate measures of dispersion around a mean effect size. They provide information about the likely effect size one can expect if we were to randomly sample a new population. For example, assume that we were interested in the overall impact of salinity stress in freshwater fish. To tackle this question, we collected experimental studies measuring fish swim performance under high salinity treatments (\~15-20ppm Na Cl) compared to freshwater (\~0ppm Na Cl) controls; using SMD to quantify the magnitude of effect salinity had on swim performance. We conduct a meta-analysis and estimate an overall SMD of 0.50 with a prediction interval of 0.05 to 0.85. The PI indicates that effects can vary widely from as low as an SMD of 0.05 to as high as an SMD 0.95, depending on the population. In addition, if we were to repeat a study we would expect a new effect to fall within the range of 0.05 and 0.85, 95% of the time [@Borenstein2019; @Nakagawa2021c]. This is a very intuitive interpretation of how variable effects are, yet PIs are often not reported in meta-analyses [@Nakagawa2021c], and we encourage more meta-analysts to report these.

@@

Comparative physiologists are interested in meta-analysing effects across a diversity of species, experimental designs and environments. Importantly, sampling variances for many effect sizes commonly used by comparative physiologists (e.g., CTmax, $Q_{10}$ etc) can be formalised and powerful meta-analytic models can be easily applied. While an overall meta-analytic mean may be of interest, more often than not, understanding the drivers of effect size heterogeneity will be the primary interest. However, factors such as temperature and dosage differences across studies may generate obvious sources of heterogeneity that may not be of prime interest.  Instead, comparative physiologists might simply want to focus on biological drivers of effect size variability and want an effect size or analytic approach that allows them to remove these ‘nuisances’. Here, we provide a set of tools to deal with ‘nuisance heterogeneity’ by ‘reinventing’ standard effect sizes and/or using mean centered nuisance variables in meta-regression models. Estimating complex meta-regression models with a multitude of fixed effects might result, however, in more challenging model interpretation. As such, a combined approach, where some of the known heterogeneity is dealt with by using an appropriate effect size and associated sampling variance while other biologically interesting moderators are fit in a meta-regression model might be desired. Indeed, this may even be a necessity to simplify the model. We wish the set of tools we describe provide clarification around importance and limitations of heterogeneity. We also hope to provide new ways to help comparative physiologists communicate effect measures more richly to guide our understanding and decisions around pressing global challenges.

@@

We would like to thank the Editorial board (Craig Franklin, Michaela Handel and Charlotte Rutledge) for inviting us to contribute to this special issue along with feedback on previous drafts. We would also like to than Megan Head for comments on a previous draft. DWAN, SN, ML and REO were supported by ARC Discovery Projects (DP210101152 to DWAN and DP200100367 to SN, ML and DWAN). PP and SB were supported by a UNSW Scientia PhD Scholarship.
